---
title: "Individual Actuarial Project"
author: "Amba Sharma"
output:
  html_document:
    df_print: paged
---

## Libraries

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
```

## Seeds for reproducibility and number of Monte Carlo simulations

```{r}
set.seed(2026)

sims <- 10000
```

# Parameters

## Japan Parameters

```{r}

jlambda <- 3.5      # Annual flood frequency (events/year)
jmu <- 9.9          # Mean economic loss per event (USD billion)
jsigma <- 4.5       # SD of economic loss (USD billion)
jp <- 0.50          # Insurance penetration rate (50%)
jr <- 0.85          # Recovery rate (85% of insured losses paid)
jgdp <- 4230        # GDP (USD billion)

```

## Bangladesh Parameters

```{r}

blambda <- 3.38     # Annual flood frequency (1.86 flood + 1.52 cyclone)
bmu <- 1.09         # Mean economic loss per event (USD billion)
bsigma <- 0.9       # SD of economic loss (USD billion)
bp <- 0.05         # Insurance penetration rate (0.05% - baseline)
br <- 0.65         # Recovery rate (65%)
bgdp <- 460         # GDP (USD billion)

```

## Bangladesh Scenario Penetration Rates

```{r}

scenarios <- data.frame(
  scenario = c("Baseline", "Low", "Medium", "High", "Japan Like"),
  penetration = c(0.05, 0.1, 0.2, 0.3, 0.5),
  description = c("Current (95% gap)", 
                  "2x current", 
                  "Regional improvement target", 
                  "Significant Policy intervention", 
                  "Matches Japan")
)

print(scenarios)
```

# Simulation functions

```{r}

simulate_annual_losses <- function(lambda, mu, sigma, sims) {
   annual_losses <- numeric(sims)
  
  for (i in 1:sims) {
    # Number of flood events this year (Poisson)
    n_events <- rpois(1, lambda)
    
    if (n_events > 0) {
      # Convert mean/sd to lognormal parameters
      # For lognormal: if X ~ LN(meanlog, sdlog), then
      # E[X] = exp(meanlog + sdlog^2/2)
      # Var[X] = exp(2*meanlog + sdlog^2)*(exp(sdlog^2) - 1)
      
      meanlog <- log(mu^2 / sqrt(sigma^2 + mu^2))
      sdlog <- sqrt(log(1 + (sigma^2 / mu^2)))
      
      # event severities
      event_losses <- rlnorm(n_events, meanlog, sdlog)
      
      # total annual loss
      annual_losses[i] <- sum(event_losses)
    } else {
      annual_losses[i] <- 0
    }
  }
  
  return(annual_losses)
}

# Test the function with Japan parameters
test_japan <- simulate_annual_losses(jlambda, jmu, jsigma, sims = 1000)

cat("Test Results (Japan, 1000 simulations):\n")
cat("Mean annual loss: $", round(mean(test_japan), 2), " billion\n", sep = "")
cat("Expected (lambda × mu): $", round(jlambda * jmu, 2), " billion\n", sep = "")
cat("SD of annual loss: $", round(sd(test_japan), 2), " billion\n")
```

# Calculating the Protection Gaps

```{r}

calculate_protection_gap <- function(annual_losses, penetration, recovery_rate) {
  # Insured loss = Economic loss × Penetration × Recovery rate
  insured_losses <- annual_losses * penetration * recovery_rate
  
  # Protection gap = Economic loss - Insured loss
  protection_gaps <- annual_losses - insured_losses
  
  # Protection gap percentage
  gap_pct <- ifelse(annual_losses > 0, 
                    protection_gaps / annual_losses * 100, 
                    0)
  
  return(list(
    economic_loss = annual_losses,
    insured_loss = insured_losses,
    protection_gap = protection_gaps,
    gap_pct = gap_pct
  ))
}
```

## Simulations

```{r}
# simulate annual losses
cat("Running simulations with n =", sims, "iterations...\n\n")

# Japan simulation
japan_losses <- simulate_annual_losses(jlambda, jmu, jsigma, sims)
japan_results <- calculate_protection_gap(japan_losses, jp, jr)

# Bangladesh simulation 
bangladesh_losses <- simulate_annual_losses(blambda, bmu, bsigma, sims)

# results for each scenario
results_list <- list()

# Japan baseline
results_list[["Japan_Baseline"]] <- data.frame(
  scenario = "Japan_Baseline",
  mean_economic_loss = mean(japan_results$economic_loss),
  mean_insured_loss = mean(japan_results$insured_loss),
  mean_protection_gap = mean(japan_results$protection_gap),
  protection_gap_pct = mean(japan_results$gap_pct),
  VaR_95 = quantile(japan_results$protection_gap, 0.95),
  VaR_99 = quantile(japan_results$protection_gap, 0.99)
)

# Bangladesh scenarios
for (i in 1:nrow(scenarios)) {
  scen_name <- paste0("Bangladesh_", scenarios$scenario[i])
  pen_rate <- scenarios$penetration[i]
  
  bg_results <- calculate_protection_gap(bangladesh_losses, pen_rate, br)
  
  results_list[[scen_name]] <- data.frame(
    scenario = scen_name,
    mean_economic_loss = mean(bg_results$economic_loss),
    mean_insured_loss = mean(bg_results$insured_loss),
    mean_protection_gap = mean(bg_results$protection_gap),
    protection_gap_pct = mean(bg_results$gap_pct),
    VaR_95 = quantile(bg_results$protection_gap, 0.95),
    VaR_99 = quantile(bg_results$protection_gap, 0.99)
  )
}

# single dataframe
model_outputs <- bind_rows(results_list)
rownames(model_outputs) <- NULL

#results
print(model_outputs, digits = 3)
```

# Visualizations

## Plot One: Protection Gap Comparison

```{r}
bangladesh_plot <- model_outputs |>
  filter(grepl("Bangladesh", scenario)) |>
  mutate(scenario_label = gsub("Bangladesh_", "", scenario))

# Reorder scenarios
bangladesh_plot$scenario_label <- factor(bangladesh_plot$scenario_label,
                                          levels = c("Baseline", "Low", "Medium", "High", "Japan Like"))

# Get Japan's protection gap for reference line
japan_gap <- model_outputs |>
  filter(scenario == "Japan_Baseline") |>
  pull(protection_gap_pct)

p1 <- ggplot(bangladesh_plot, aes(x = scenario_label, y = protection_gap_pct)) +
  geom_bar(stat = "identity", fill = "lightblue", width = 0.6) +
  geom_hline(yintercept = japan_gap, linetype = "dashed", color = "darkblue", linewidth = 1.2) +
  geom_text(aes(label = paste0(round(protection_gap_pct, 1), "%")), 
            vjust = -0.5, size = 4) +
  annotate("text", x = 1.5, y = japan_gap - 5, 
           label = paste0("Japan: ", round(japan_gap, 1), "%"),
           color = "darkblue", fontface = "bold", size = 4) +
  labs(
    title = "Bangladesh Protection Gap Under Different Scenarios",
    subtitle = "Dashed blue line shows Japan's current protection gap for comparison",
    x = "Insurance Penetration Scenario",
    y = "Protection Gap (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    axis.text.x = element_text(size = 10)
  ) +
  ylim(0, 105)

print(p1)
```

## Plot 2: VaR Comparisons (Tail Risk)

```{r}

var_plot_data <- model_outputs |>
  filter(grepl("Bangladesh", scenario)) |>
  mutate(scenario_label = gsub("Bangladesh_", "", scenario)) |>
  select(scenario_label, VaR_95, VaR_99) |>
  tidyr::pivot_longer(cols = c(VaR_95, VaR_99), 
                      names_to = "VaR_level", 
                      values_to = "value")

# Reorder scenarios
var_plot_data$scenario_label <- factor(var_plot_data$scenario_label,
                                        levels = c("Baseline", "Low", "Medium", "High", "Japan Like"))

# Clean up labels
var_plot_data$VaR_level <- gsub("VaR_", "VaR ", var_plot_data$VaR_level)
var_plot_data$VaR_level <- paste0(var_plot_data$VaR_level, "%")

# Get Japan VaR values for reference lines
japan_var95 <- model_outputs |> filter(scenario == "Japan_Baseline") |> pull(VaR_95)
japan_var99 <- model_outputs |> filter(scenario == "Japan_Baseline") |> pull(VaR_99)

p2 <- ggplot(var_plot_data, aes(x = scenario_label, y = value, fill = VaR_level)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = paste0("$", round(value, 1), "bn")), 
            position = position_dodge(width = 0.7), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("VaR 95%" = "lightblue", "VaR 99%" = "darkblue")) +
  labs(
    title = "Uninsured Losses in Extreme Years: Bangladesh Scenarios",
    subtitle = "VaR 95% = 1-in-20 year loss | VaR 99% = 1-in-100 year loss",
    x = "Insurance Penetration Scenario",
    y = "Uninsured Loss (USD Billion)",
    fill = "Risk Measure"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "top"
  ) +
  ylim(0, 14)

print(p2)
```

## Plot 3: GDP-Relative Impact Chart

```{r}

# Japan's GDP
gdp_plot <- model_outputs |>
  mutate(
    gdp = ifelse(scenario == "Japan_Baseline", 4230, 460),
    gap_pct_gdp = (mean_protection_gap / gdp) * 100
  )

# comparison dataframe
gdp_comparison <- data.frame(
  country = c("Japan", "Bangladesh\n(Current)", "Bangladesh\n(Japan-Like)"),
  gap_pct_gdp = c(
    gdp_plot$gap_pct_gdp[gdp_plot$scenario == "Japan_Baseline"],
    gdp_plot$gap_pct_gdp[gdp_plot$scenario == "Bangladesh_Baseline"],
    gdp_plot$gap_pct_gdp[gdp_plot$scenario == "Bangladesh_Japan Like"]
  )
)

gdp_comparison$country <- factor(gdp_comparison$country, 
                                  levels = c("Japan", "Bangladesh\n(Current)", "Bangladesh\n(Japan-Like)"))

p3 <- ggplot(gdp_comparison, aes(x = country, y = gap_pct_gdp, fill = country)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = paste0(round(gap_pct_gdp, 2), "%")), vjust = -0.5, size = 4.5) +
  scale_fill_manual(values = c("Japan" = "lightblue", 
                                "Bangladesh\n(Current)" = "#3498DB",
                                "Bangladesh\n(Japan-Like)" = "darkblue")) +
  labs(
    title = "Annual Uninsured Flood Losses as Percentage of GDP",
    subtitle = "Economic burden relative to national income",
    x = "",
    y = "Uninsured Loss (% of GDP)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "none"
  ) +
  ylim(0, 1)

print(p3)

```

## Plot 4: Distribution of simulated annual losses

```{r}

hist(bangladesh_losses, breaks = 50, main = "Distribution of Simulated Annual Losses (Bangladesh)",
     xlab = "Annual Loss (USD Billion)", col = "lightblue", border = "white")
abline(v = mean(bangladesh_losses), col = "darkblue", lwd = 2, lty = 2)
```
Loss Distribution Validation Against Observed Data
```{r}

# Actual Bangladesh event losses from calibration data (USD billion)
observed_bd <- c(0.032, 0.13, 0.50, 0.80, 1.30, 1.50, 1.70, 2.80)

# Simulate individual event losses using same lognormal parameters
meanlog_bd <- log(bmu^2 / sqrt(bsigma^2 + bmu^2))
sdlog_bd   <- sqrt(log(1 + bsigma^2 / bmu^2))

set.seed(2026)
sim_events <- rlnorm(10000, meanlog = meanlog_bd, sdlog = sdlog_bd)

# Calculate percentile rank of each observed event
cat("Validation: Observed events vs simulated event distribution\n\n")
cat(sprintf("%-25s %15s %15s\n", "Event", "Loss (USD bn)", "Percentile"))
cat(paste(rep("-", 55), collapse = ""), "\n")

for (loss in sort(observed_bd)) {
  pctile <- round(mean(sim_events <= loss) * 100, 1)
  cat(sprintf("%-25s %15.3f %14.1f%%\n", "", loss, pctile))
}

cat("\nSimulated event distribution summary:\n")
cat("  Mean:   USD", round(mean(sim_events), 2), "bn\n")
cat("  Median: USD", round(median(sim_events), 2), "bn\n")
cat("  SD:     USD", round(sd(sim_events), 2), "bn\n")
cat("  Min:    USD", round(min(sim_events), 4), "bn\n")
cat("  Max:    USD", round(max(sim_events), 2), "bn\n")

# Histogram with observed events overlaid
hist(sim_events[sim_events < 10], breaks = 100,
     main = "Simulated vs Observed Individual Event Losses (Bangladesh)",
     xlab = "Event Loss (USD Billion)", col = "lightblue", border = "white",
     xlim = c(0, 10))
abline(v = observed_bd, col = "red", lwd = 2, lty = 2)
legend("topright",
       legend = c("Simulated distribution", "Observed events"),
       col = c("lightblue", "red"),
       lwd = c(10, 2), lty = c(1, 2))
```

# Sensitivity Analysis

## Recovery rate sensitivity - penetration fixed at 5%

```{r}
recovery_test <- c(0.55, 0.65, 0.75)
penetration_baseline <- 0.05

for (r in recovery_test) {
  gap_pct <- (1 - penetration_baseline * r) * 100
  cat("Recovery rate", r, "-> Protection gap:", round(gap_pct, 1), "%\n")
}
```

## Penetration Rate Sensitivity - recovery fixed at 65%

```{r}
penetration_test <- c(0.05, 0.10, 0.15)
recovery_baseline <- 0.65

for (p in penetration_test) {
  gap_pct <- (1 - p * recovery_baseline) * 100
  cat("Penetration rate", p, "-> Protection gap:", round(gap_pct, 1), "%\n")
}
```

## Tail Sensitivity - Effect of increasing sigma by 10%

```{r}

bsigma2 <- 0.99          # original 0.90 increased by 10%

# Lognormal moment-matching with NEW sigma
mu_log    <- log(bmu^2 / sqrt(bsigma2^2 + bmu^2))
sigma_log <- sqrt(log(1 + bsigma2^2 / bmu^2))

# Simulate aggregate annual losses
agg_losses <- numeric(sims)
for (i in 1:sims) {
  n_events <- rpois(1, blambda)
  if (n_events > 0) {
    event_losses <- rlnorm(n_events, meanlog = mu_log, sdlog = sigma_log)
    agg_losses[i] <- sum(event_losses)
  } else {
    agg_losses[i] <- 0
  }
}

# Calculate protection gap for each simulated year
insured    <- agg_losses * bp * br
prot_gap   <- agg_losses - insured

# VaR 99% of the protection gap
var99_sens <- quantile(prot_gap, 0.99)

cat("\nTail Sensitivity (sigma +10%):\n")
cat("VaR 99% protection gap: USD", round(var99_sens, 2), "billion\n")
```

# Summary Statistics

```{r}
cat("KEY FIGURES FOR REPORT \n\n")
cat("Simulation: 10,000 iterations\n")
cat("Bangladesh flood frequency: λ =", blambda, "events/year\n")
cat("Bangladesh mean severity: μ = $", bmu, "bn\n\n")
cat("BASELINE COMPARISON:\n")
cat("Japan protection gap: 55.9%\n")
cat("Bangladesh protection gap:", 
    round(model_outputs$protection_gap_pct[model_outputs$scenario == "Bangladesh_Baseline"], 1), "%\n")
cat("SCENARIO IMPACT:\n")
cat("Bangladesh at 50% penetration: 67.5% gap\n")
cat("Gap reduction from baseline: 27.1 percentage points\n\n")
cat("GDP-RELATIVE BURDEN:\n")
cat("Japan: 0.48% of GDP\n")
cat("Bangladesh current: 0.74% of GDP\n")
cat("Bangladesh Japan-like: 0.51% of GDP\n")
```
